p-adic stability in linear algebra

### 2015-01-15 examples/stability.tex
Mostly spelling and grammar changes

\newcommand{\detp}{\det'}
\newcommand{\low}{\text{\rm low}}
\newcommand{\up}{\text{\rm up}}
\newcommand{\DI}{\text{\rm DI}}
\newcommand{\II}{\text{\rm II}}
\DeclareMathOperator{\charpoly}{char}
\newcommand{\charp}{\charpoly'}

@@

For about twenty years, the use of $p$-adic methods in symbolic 
computation was gaining more and more popularity: they were succesfully 
used for instance 
to compute compose products of polynomials 
\cite{boston-gonzalez-perdry-schost:05a}, 
to produce hyperelliptic curves of genus $2$ with complex multiplication 
\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
and, last but not least,
to count points on certain varieties (mainly curves) using $p$-adic 
cohomologies (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
followers).
However, a general framework allowing a precise study of $p$-adic 
precision --- which is a main issue we encounter when trying to work 
with $p$-adic numbers --- was designed only recently in 
\cite{caruso-roe-vaccon:14a}.

@@

\begin{comment}
The present paper, which may be considered as a continuation of 
\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. 
cit.} in order to analyse the $p$-adic stability of many standard 
algorithms for computing with basic algebraic structures: numbers, 
matrices, univariate polynomials and vector spaces. For each considered 
algorithm, we will follow the same protocol: 
first, we will compute the theoretical loss of precision of the 
underlying question using the framework of \cite{caruso-roe-vaccon:14a} 
and then compare it to those observed in available implementations in 
\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
\textsc{sage} \cite{sage}.

@@

\noindent
{\bf Organization of the paper.}
The section \ref{sec:theory} introduces the theory of ultrametric 
precision developped in \cite{caruso-roe-vaccon:14a} and completes 
it on a technical point.
In \S \ref{sec:matrices}, we focus on matrices... \todo{Complete
this}.

@@

\noindent
{\bf Notations.}
Throughout the paper, the letter $K$ will refer to a \emph{complete 
discrete valuation field of characteristic $0$}, that is either a finite 
extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
series over a field of characteristic $0$. We denote by $\val : K \to \Z 
\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
$\Vert \cdot \Vert$ the corresponding norm.

@@

The aim of this section is to summerize briefly the content of 
\cite{caruso-roe-vaccon:14a} and complete it on certain points.

@@

In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be made explicit in terms of the
growing function $\Lambda(f)$ of $f$ defined by:
$$\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)$$
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement and seek here to state the particular case of
integral polynomial functions. A function $f : E \to F$ is said 
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to orthonormal basis.

@@

We actually generalize a bit the setting above and consider the 
differential equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
normed vector spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a normed $K$-vector space $G$. Consequently,
$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
all (continuous) linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$ and second, in order to 
apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
needs to be surjective and therefore \emph{a fortiori} nonzero.

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert > 0$ where $p$ denotes the 
characteristic of the residue field and where, by convention, the above 
expression equals $0$ if $p = 0$.
The next Proposition is proved in Appendix \ref{app:proof}.

\begin{prop}
\label{prop:boundLambdaf}\label{PROP:BOUNDLAMBDAF}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

@@

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$]{-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(x)$ starts with the term
$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

@@

\begin{prop}
\label{prop:boundLambdaf2}
Keeping the above notations, we have:
$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$.
\end{prop}

@@

\begin{rem}
In certain situations,
it may happen that the function $f$ is solution of a simplier 
differential equation of the form $f' = g \circ f$. If this holds, 
Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.

Beyond this particular case, our general recommandation for choosing the 
function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
underlying vector spaces. The function $\Lambda(h)$ the maps $x$ to $x + 
\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
\end{rem}

@@

To begin with, we want to study the behaviour of the precision when 
performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
positive integers and assume that we want to multiply a matrix $A \in 
M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
modeled by the integral polynomial function:
$$\begin{array}{rcl}
\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
M_{r,t}(K) \smallskip \\
(A,B) & \mapsto & AB.
\end{array}$$
According to Proposition \ref{prop:precision}, the behaviour of the precision when 
computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
that this differential is the linear mapping that takes a couple 
$(dA,dB)$ to $A \cdot dB + dA \cdot B$.

@@

\begin{prop}
\label{prop:mulmatrix}
With the above notations, we have:
$$\begin{array}{rl}
& \mathcal P'_{a,b,c}(A,B)(\mathcal L_0)
= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
\text{and} &
\text{\emph{length}}\Big(\frac{\mathcal P'_{a,b,c}(A,B)(\mathcal L_0) }
{M_{r,t}(\O_K)}\Big) =
\displaystyle 
\sum_{m=1}^\infty a^{(m)} b^{(m)}
\end{array}$$
where $a^{(m)}$ (resp. $b^{(m)}$) is the number of $a_i$'s (resp.
of $b_j$'s) which are $\geq m$.
\end{prop}

@@

One may wonder to what extend it is important to take care of these 
``diffused'' gains of precision. For one single product of matrices, 
it is not so much. However, the benefits becomes really substantial
when we are executing an algorithm that computes many products of
matrices. Let us illustrate this with the following quite simple
example.

@@

\noindent
Figure \ref{fig:mulmatrix} compares the number of significant digits in 
\emph{relative} precision we are loosing on the output of Algorithm 1 
when we are using, on the one hand, a standard coordinate-wise track of 
precision and, on the other hand, a lattice-based method to handle 
precision.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{2}{|c|}{Average loss of precision} \\
\cline{3-4}
& & Coord-wise method & Lattice method \\
\hline 
$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
$2$ & $1000$ & $157.8$ & $7.9$ \\
\hline
$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
$3$ & $1000$ & $122.5$ & $7.0$ \\
\hline
\end{tabular}

@@

We now come to the computation of the determinant of the square $n 
\times n$ with coefficients in $K$. The differential computation is 
quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
K$ at some point $M$ is the linear mapping
$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),$$
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.

@@

\begin{proof}
From the description of $\detp(M)$, we see that it is enough to prove 
that the smallest valuation on an entry of $\com(M)$ is $v$
or, equivalently, that the ideal of $\O_K$ generated by all 
minors of $M$ of size $(n-1)$ is $\pi^v
\O_K$. But this ideal remains unchanged when we multiply $M$ 
(on the left or on the right) by any unimodular matrix. Thus we may 
assume that $M$ is $\Delta_M$ and the result becomes clear.
\end{proof}

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$, that is a writing
$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and has diagonal entries $a_1, \ldots, a_m$ of valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.  In the standard case that the precision lattice for $M$ is just $H = \pi^k M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.

@@

If a polygon $P$ has vertices $(x_i, y_i)$, we $T_n(P)$ to be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.

\begin{prop}
Suppose a matrix $M$ is known with precision $H = \pi^k M_{n,n}(\O_K)$.  Then the precision of the characteristic polynomial of $M$ is contained within $S_k$ applied to the lattice of the precision polygon of $M$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
\end{proof}

@@

\noindent $\bullet$
$L_n$ (resp. $L_n^u$) is the sub-affine space space of 
nilpotent (resp. unipotent) lower-triangular matrices.

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$val(\det(D'(M)))$, which is the lattice-method loss of precision.
Figure \ref{fig:LU} sums up the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_d(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip

>>>>

\newcommand{\detp}{\det{'}}
\newcommand{\low}{\text{\rm low}}
\newcommand{\up}{\text{\rm up}}
\newcommand{\DI}{\text{\rm DI}}
\newcommand{\II}{\text{\rm II}}
\DeclareMathOperator{\charpoly}{char}
\newcommand{\charp}{\charpoly'}

@@

For about twenty years, the use of $p$-adic methods in symbolic 
computation has been gaining popularity: they were successfully 
used for instance 
to compute compose products of polynomials 
\cite{boston-gonzalez-perdry-schost:05a}, 
to produce hyperelliptic curves of genus $2$ with complex multiplication 
\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
and, last but not least,
to count points on certain varieties (mainly curves) using $p$-adic 
cohomology theories (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
followers).
However, a general framework allowing a precise study of $p$-adic 
precision --- which is a main issue we encounter when trying to work 
with $p$-adic numbers --- was designed only recently in 
\cite{caruso-roe-vaccon:14a}.

@@

\begin{comment}
The present paper, which may be considered as a continuation of 
\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. 
cit.} in order to analyze the $p$-adic stability of many standard 
algorithms for computing with basic algebraic structures: numbers, 
matrices, univariate polynomials and vector spaces. For each considered 
algorithm, we will follow the same protocol: 
first, we will compute the theoretical loss of precision of the 
underlying question using the framework of \cite{caruso-roe-vaccon:14a} 
and then compare it to those observed in available implementations in 
\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
\textsc{sage} \cite{sage}.

@@

\noindent
{\bf Organization of the paper.}
The section \ref{sec:theory} introduces the theory of ultrametric 
precision developed in \cite{caruso-roe-vaccon:14a} and completes 
it on a technical point.
In \S \ref{sec:matrices}, we focus on matrices... \todo{Complete
this}.

@@

\noindent
{\bf Notations.}
Throughout the paper, the letter $K$ will refer to a \emph{complete 
discrete valuation field of characteristic $0$}, \todo{why characteristic 0?} that is either a finite 
extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
series over a field of characteristic $0$. We denote by $\val : K \to \Z 
\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
$\Vert \cdot \Vert$ the corresponding norm.

@@

The aim of this section is to summarize briefly the content of 
\cite{caruso-roe-vaccon:14a} and complete it on certain points.

@@

In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be made explicit in terms of the
growing function $\Lambda(f)$ of $f$ defined by:
$$\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)$$
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement and seek here to state the particular case of
integral polynomial functions. A function $f : E \to F$ is said 
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to orthonormal bases.

@@

We actually generalize a bit the setting above and consider the 
differential equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
normed vector spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a normed $K$-vector space $G$. Consequently,
$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ of
(continuous) linear maps $E \to F$.
In what follows, we always assume that $V$ and $W$ contain the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$ and second, in order to 
apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
needs to be surjective and therefore \emph{a fortiori} nonzero.

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fullfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert > 0$ where $p$ denotes the 
characteristic of the residue field and where, by convention, the above 
expression equals $0$ if $p = 0$. \todo{which expression; the one defining $\alpha$?}
The next Proposition is proved in Appendix \ref{app:proof}.

\begin{prop}
\label{prop:boundLambdaf}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

@@

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue \todo{do we have color?} plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(z)$ starts with the term
$g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

@@

\begin{prop}
\label{prop:boundLambdaf2}
Keeping the above notations, we have:
$$\Lambda(f)_{\leq 2} (x) \leq 2(x + \alpha + \mu) -
\min(\Lambda_h^{-1}(\nu) + \mu, \: \nu)$$
for all $x \leq \min(\Lambda_h^{-1}(\nu) - \alpha, \: \nu - \mu - \alpha)$. \todo{What is $\Lambda(f)_{\leq 2}$?}
\end{prop}

@@

\begin{rem}
In certain situations,
it may happen that the function $f$ is solution of a simpler 
differential equation of the form $f' = g \circ f$. If this holds, 
Proposition \ref{prop:boundLambdaf2} gives the bound $\Lambda(f)_{\leq 
2} (x) \leq 2(x + \alpha + \mu) - \nu$ for $x \leq \nu - \mu - \alpha$.

Beyond this particular case, our general recommendation for choosing the 
function $h$ is to endow $F$ with the second norm $\Vert x \Vert'_F = 
e^\mu \cdot \Vert x \Vert_F$ ($x \in F$) and to take $h : (F, \Vert 
\cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ acting as identity on the
underlying vector spaces. The function $\Lambda(h)$ then maps $x$ to $x + 
\mu$ ($x \in \R$) and we can choose $\Lambda_h = \Lambda(h)$.
\end{rem}

@@

To begin with, we want to study the behavior of the precision when 
performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
positive integers and assume that we want to multiply a matrix $A \in 
M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
modeled by the integral polynomial function:
$$\begin{array}{rcl}
\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
M_{r,t}(K) \smallskip \\
(A,B) & \mapsto & AB.
\end{array}$$
According to Proposition \ref{prop:precision}, the behavior of the precision when 
computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$, the linear mapping that takes a pair 
$(dA,dB)$ to $A \cdot dB + dA \cdot B$.

@@

\begin{prop}
\label{prop:mulmatrix}
With the above notations, we have:
$$\begin{array}{rl}
& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
\text{and} &
\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
\Big) =
\displaystyle 
\sum_{m=1}^\infty a^{(m)} b^{(m)}
\end{array}$$
where $a^{(m)}$ (resp. $b^{(m)}$) is the number of $a_i$'s (resp.
of $b_j$'s) which are $\geq m$.
\end{prop}

@@

One may wonder to what extent it is important to take care of these 
``diffused'' gains of precision. For one single product of matrices, 
it is not so much. However, the benefits become substantial
when we are executing an algorithm that computes many products of
matrices. Let us illustrate this with the following quite simple
example.

@@

\noindent
Figure \ref{fig:mulmatrix} compares the number of significant digits in 
\emph{relative} precision we are losing on the output of Algorithm 1 
when we are using, on the one hand, a standard coordinate-wise track of 
precision and, on the other hand, a lattice-based method to handle 
precision.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{2}{|c|}{Average loss of precision} \\
\cline{3-4}
& & Coord-wise method & Lattice method \\
\hline 
$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
$2$ & $1000$ & $157.8$ & $7.9$ \\
\hline
$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
$3$ & $1000$ & $122.5$ & $7.0$ \\
\hline
\end{tabular}

@@

We now come to the computation of the determinant of a square $n 
\times n$ matrix with coefficients in $K$. The differential computation is 
quite classical.  The differential of the function $\det : M_{n,n}(K) \to 
K$ at some point $M$ is the linear mapping
$$\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),$$
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.

@@

\begin{proof}
From the description of $\detp(M)$, we see that it is enough to prove 
that the smallest valuation of an entry of $\com(M)$ is $v$
or, equivalently, that the ideal of $\O_K$ generated by all 
minors of $M$ of size $(n-1)$ is $\pi^v
\O_K$. But this ideal remains unchanged when we multiply $M$ 
(on the left or on the right) by any unimodular matrix. Thus we may 
assume that $M$ is $\Delta_M$ and the result becomes clear.
\end{proof}

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$, that is a writing
$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and has diagonal entries $a_1, \ldots, a_n$ of valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

@@

For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.  In the standard case that the precision lattice for $M$ is just $H = \pi^N M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.

@@

If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.

\begin{prop}
Suppose a matrix $M$ is known with precision $H = \pi^N M_{n,n}(\O_K)$.  Then the precision of the characteristic polynomial of $M$ is contained within $S_N$ applied to the lattice of the precision polygon of $M$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
\end{proof}

@@

\noindent $\bullet$
$L_n$ (resp. $L_n^u$) is the sub-affine space of 
nilpotent (resp. unipotent) lower-triangular matrices.

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$\val(\det(D'(M)))$, which is the lattice-method loss of precision.
Figure \ref{fig:LU} sums up the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_d(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip
